---
title: "R Notebook"
# output: html_notebook
---

Poa INLA model - this is the script testing the influence of endophytes on range size. This goes over different phylogenetic priors, meshes, and spatial priors. The exact same models were also run for our other niche breadth variables: range on the primary niche axis and hypervolume. 

These models were also used to test how environmental variables structure endophyte presence - this is the same script (tested the same priors and meshes) but those models use a binomial family. 



setup chunk
```{r setup}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```


```{r}
library(INLA)
library(ape)
library(sf)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(viridis)
```


load in data
```{r}
trait_data <- read.csv("data/processed/Poa_data.csv")
occ_data <- read.csv("data/processed/Poa_occurrence_records.csv")
phy <- read.tree("data/processed/Poa_tree.nex")
```


have a look at the occurrence data - I am having to rewrite this script to use terra
```{r}
#load in background map for plotting
coast <- read_sf("data/raw/ne_50m_land.shp")

ggplot() + geom_sf(data = coast, colour = "black", fill = NA) + geom_point(aes(x=decimalLongitude, y=decimalLatitude), occ_data, size=0.1, colour= "red")
```



Match all the data to the tree:
```{r}
rownames(trait_data) <- trait_data$WCVP_list

matched_tree_df <- picante::match.phylo.data(phy, trait_data)

Poa_tree <- matched_tree_df$phy
Poa_data_reduced <- matched_tree_df$data

Poa_data_reduced$Endophyte_status <- as.integer(Poa_data_reduced$Endophyte_status)

Poa_data_reduced$log_maxss.range.size.sqkm <- as.numeric(Poa_data_reduced$log_maxss.range.size.sqkm)
Poa_data_reduced$log_top.var.range <- as.numeric(Poa_data_reduced$log_top.var.range)
Poa_data_reduced$log_volume <- as.numeric(Poa_data_reduced$log_volume)
Poa_data_reduced$centroid_latitude <- as.numeric(Poa_data_reduced$centroid_latitude)


Poa_data_reduced$bio1 <- as.numeric(Poa_data_reduced$bio1)
Poa_data_reduced$bio12 <- as.numeric(Poa_data_reduced$bio12)
Poa_data_reduced$nitrogen_0_30cm_mean <- as.numeric(Poa_data_reduced$nitrogen_0_30cm_mean)
Poa_data_reduced$clay_0_5cm_mean_1000 <- as.numeric(Poa_data_reduced$clay_0_5cm_mean_1000)
```


```{r}
poa_occ <- occ_data %>%
   dplyr::select(species, Long = decimalLongitude, Lat = decimalLatitude) %>% 
   ## make sure species names use same format as traits 
   transform(species = gsub(" ", "_", species)) 

species_to_remove <- setdiff(unique(poa_occ$species), unique(phy$tip.label))

poa_occ_filtered <- poa_occ[!poa_occ$species %in% species_to_remove, ]


str(poa_occ)
str(poa_occ_filtered)
```


```{r}
poa_traits <- Poa_data_reduced %>%
  dplyr::select(species = WCVP_list, Endophyte_status, log_maxss.range.size.sqkm,log_top.var.range,log_volume, centroid_latitude, bio1, bio12, nitrogen_0_30cm_mean, clay_0_5cm_mean_1000)


poa_traits<- poa_traits %>% 
  mutate(centroid_latitude_abs = abs(centroid_latitude),
         log_maxss.range.size.sqkm_st = (log_maxss.range.size.sqkm - mean(log_maxss.range.size.sqkm)) / sd(log_maxss.range.size.sqkm),
         log_top.var.range_st = (log_top.var.range - mean(log_top.var.range)) / sd(log_top.var.range),
         log_volume_st = (log_volume - mean(log_volume)) / sd(log_volume),
         centroid_latitude_st = (centroid_latitude - mean(centroid_latitude)) / sd(centroid_latitude),
         bio1_st = (bio1 - mean(bio1)) / sd(bio1), 
         bio12_st = (bio12 - mean(bio12)) / sd(bio12), 
         nitrogen_0_30cm_mean_st = (nitrogen_0_30cm_mean - mean(nitrogen_0_30cm_mean)) / sd(nitrogen_0_30cm_mean),
         clay_0_5cm_mean_1000_st = (clay_0_5cm_mean_1000 - mean(clay_0_5cm_mean_1000)) / sd(clay_0_5cm_mean_1000))


poa_traits_occ <- left_join(poa_occ_filtered, poa_traits, by = "species")

poa_traits_occ
```


Creating the phylogenetic precision matrix to use for the phylogenetic effect
```{r}
phylo_covar_mat <- ape::vcv(phy)
phylo_covar_mat <- phylo_covar_mat/max(phylo_covar_mat)
phylo_covar_mat <- phylo_covar_mat / exp(determinant(phylo_covar_mat)$modulus[1] /
                                           nrow(phylo_covar_mat))

phylo_prec_mat <- solve(phylo_covar_mat) 
phylo_prec_mat[1:5, 1:5] 
```


```{r}
poa_traits <- poa_traits %>%
  left_join(tibble(species = rownames(phylo_prec_mat), 
                   phy_id = 1:nrow(phylo_prec_mat)))
```




Setting the phylogenetic prior, will try a range of priors, first using a default prior
```{r}
pcprior <- list(prec = list(prior="pc.prec"))
```


Will first run an INLA model just with the phylogenetic effect
```{r}
Endophyte_model_phylo <- inla(log_maxss.range.size.sqkm_st ~ Endophyte_status +
                                f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                              family = "gaussian",
                              data = poa_traits,
                              control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)

summary(Endophyte_model_phylo)
```




Testing other phylogentic priors
```{r}
pcprior <- list(prec = list(prior="pc.prec", param = c(0.5, 0.05)))
```

```{r}
Endophyte_model_phylo <- inla(log_maxss.range.size.sqkm_st ~ Endophyte_status +
                                f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                              family = "gaussian",
                              data = poa_traits,
                              control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)

summary(Endophyte_model_phylo)
```




```{r}
pcprior <- list(prec = list(prior="pc.prec", param = c(1, 0.1)))
```

```{r}
Endophyte_model_phylo <- inla(log_maxss.range.size.sqkm_st ~ Endophyte_status +
                                f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                              family = "gaussian",
                              data = poa_traits,
                              control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)

summary(Endophyte_model_phylo)
```





```{r}
pcprior <- list(prec = list(prior="pc.prec", param = c(1, 0.01)))
```

```{r}
Endophyte_model_phylo <- inla(log_maxss.range.size.sqkm_st ~ Endophyte_status +
                                f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                              family = "gaussian",
                              data = poa_traits,
                              control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)

summary(Endophyte_model_phylo)
```






```{r}
pcprior <- list(prec = list(prior="pc.prec", param = c(1.5, 0.05)))
```

```{r}
Endophyte_model_phylo <- inla(log_maxss.range.size.sqkm_st ~ Endophyte_status +
                                f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                              family = "gaussian",
                              data = poa_traits,
                              control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)

summary(Endophyte_model_phylo)
```

 


```{r}
pcprior <- list(prec = list(prior="pc.prec", param = c(1, 0.05)))
```

```{r}
Endophyte_model_phylo <- inla(log_maxss.range.size.sqkm_st ~ Endophyte_status +
                                f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                              family = "gaussian",
                              data = poa_traits,
                              control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE)
)

summary(Endophyte_model_phylo)
``` 
 
 
 
 
 


Now to add in the spatial effect

First, creating a mesh - will try a few:



Mesh1
```{r}
coords <- st_as_sf(occ_data, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)


boundary <- list(inla.nonconvex.hull(coords, 5, resolution=c(60,60), crs=4326),
                 inla.nonconvex.hull(coords, 12, resolution=c(50,50), crs=4326 ))

```

```{r}
## Build the mesh:
spatial_mesh1 <- inla.mesh.2d(boundary=boundary, 
                             max.edge=c(5, 12),
                             # min.angle=30,
                             cutoff= 3, ## Filter away adjacent points. 
                             offset=c(1, 2))



plot(spatial_mesh1)
plot(spatial_mesh1); points(occ_data$decimalLongitude, occ_data$decimalLatitude, col="red", cex = 0.1)
spatial_mesh1$n
```

setting the priors for the spatial effect - just using default prior for now
```{r}
spde <- inla.spde2.matern(mesh = spatial_mesh1, alpha = 2)
```


```{r}
A <- inla.spde.make.A(spatial_mesh1, loc = coords)
dim(A)

row_counts <- rowSums(A != 0); head(row_counts)
row_sums <- rowSums(A); head(row_sums)
apply(A, 1, function(x) x[x > 0]) %>% t %>% head
```


```{r}
specs <- poa_traits$species

# Loop over species and compute one A-row per species
spec_A_list <- lapply(specs, function(x) {
  rows_x <- which(poa_traits_occ$species == x)
  A_x <- A[rows_x, , drop = FALSE]
  A_x_mean <- apply(A_x, 2, mean)

  return(A_x_mean)
})


# Combine species rows into a single matrix
spec_A_rows <- do.call(rbind, spec_A_list)

## convert back to sparse matrix
spec_A_rows <- Matrix(spec_A_rows, sparse = TRUE) 

dim(spec_A_rows)
apply(spec_A_rows, 1, sum) %>% head

apply(spec_A_rows, 1, function(x) sum(x > 0)) 

```


```{r}
stack <- inla.stack(data = list(resp = poa_traits$log_maxss.range.size.sqkm_st),
                    A = list(spec_A_rows, 1, 1, 1), 
                    effects=list(spat_id = 1:spde$n.spde,
                                 phy_id = poa_traits$phy_id,
                                 intercept = rep(1, nrow(poa_traits)), 
                                 list(Endophyte = poa_traits$Endophyte_status
                                      )), tag='est') 
```


```{r}
## estimate spatial random effect for each species
stck.loc <- inla.stack(data = list(resp = NA),
                       A = list(spec_A_rows),
                       effects = list(spat_id = 1:spde$n.spde), tag='loc') 

## estimate phylogenetic random effect for each species
stck.phy <- inla.stack(data = list(resp = NA), A = list(1),
                       effects=list(phy_id = poa_traits$phy_id), tag='phy') 

big_stack <- inla.stack(stack, stck.loc, stck.phy)
```


```{r}
full_mod <- inla(resp ~ 0 + intercept + Endophyte +
                   f(spat_id, model = spde) + ## spatial random effect
                   f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                 family = "gaussian",
                 data = inla.stack.data(big_stack, spde = spde),
                 control.predictor = list(A = inla.stack.A(big_stack), compute = TRUE,
                                          link = 1), 
                 control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE),
                 num.threads = 10 
)

summary(full_mod)
```


```{r}
grid <- inla.mesh.projector(spatial_mesh, dims = c(501,501))


prd_mean <- inla.mesh.project(grid, full_mod$summary.ran$spat_id$mean) 
prd_sd <- inla.mesh.project(grid, full_mod$summary.ran$spat_id$sd)

grid_df <- tibble(long = grid$lattice$loc[ , 1], 
                  lat = grid$lattice$loc[ , 2],
                  prediction_mean = as.vector(prd_mean), 
                  prediction_sd = as.vector(prd_sd))
```

plot
```{r}
bbox <- st_bbox(coast, crs = "+proj=longlat +datum=WGS84 +no_defs")

# Convert to sf polygon
grid_sf <- st_as_sf(grid_df, coords = c("long", "lat"), crs = 4326)

land <- st_union(coast)

grid_crop <- grid_sf[st_intersects(grid_sf, land, sparse = FALSE), ]

grid_crop <- grid_crop %>%
  mutate(long = st_coordinates(grid_crop)[,1],
         lat  = st_coordinates(grid_crop)[,2])

spat_map <- ggplot(coast) +
  geom_sf() +
  geom_raster(data = grid_crop, aes(x = long, y = lat, fill = prediction_mean)) + 
  # geom_sf(data = sea, fill = "white") +
  scale_fill_viridis(name = "Spatial effect", na.value = NA) + 
  theme_map() 

spat_map
```







Mesh 2

```{r}
coords <- st_as_sf(occ_data, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)


boundary <- list(inla.nonconvex.hull(coords, 4, resolution=c(100,100), crs=4326),
                 inla.nonconvex.hull(coords, 8, resolution=c(80,80), crs=4326 ))

```

```{r}
## Build the mesh:
spatial_mesh2 <- inla.mesh.2d(boundary=boundary, 
                             max.edge=c(2, 10),
                             # min.angle=30,
                             cutoff= 2, ## Filter away adjacent points. 
                             offset=c(1, 2)
                             )



plot(spatial_mesh2)
plot(spatial_mesh2); points(occ_data$decimalLongitude, occ_data$decimalLatitude, col="red", cex = 0.1)
spatial_mesh2$n
```

setting the priors for the spatial effect
```{r}
spde <- inla.spde2.matern(mesh = spatial_mesh2, alpha = 2)
```


```{r}
A <- inla.spde.make.A(spatial_mesh2, loc = coords)
dim(A)

row_counts <- rowSums(A != 0); head(row_counts)
row_sums <- rowSums(A); head(row_sums)
apply(A, 1, function(x) x[x > 0]) %>% t %>% head
```


```{r}
specs <- poa_traits$species

# Loop over species and compute one A-row per species
spec_A_list <- lapply(specs, function(x) {
  rows_x <- which(poa_traits_occ$species == x)
  A_x <- A[rows_x, , drop = FALSE]
  A_x_mean <- apply(A_x, 2, mean)

  return(A_x_mean)
})


# Combine species rows into a single matrix
spec_A_rows <- do.call(rbind, spec_A_list)

## convert back to sparse matrix
spec_A_rows <- Matrix(spec_A_rows, sparse = TRUE) 

dim(spec_A_rows)
apply(spec_A_rows, 1, sum) %>% head

apply(spec_A_rows, 1, function(x) sum(x > 0)) 

```


```{r}
stack <- inla.stack(data = list(resp = poa_traits$log_maxss.range.size.sqkm_st),
                    A = list(spec_A_rows, 1, 1, 1), 
                    effects=list(spat_id = 1:spde$n.spde,
                                 phy_id = poa_traits$phy_id,
                                 intercept = rep(1, nrow(poa_traits)), 
                                 list(Endophyte = poa_traits$Endophyte_status
                                      )), tag='est') 
```


```{r}
## estimate spatial random effect for each species
stck.loc <- inla.stack(data = list(resp = NA),
                       A = list(spec_A_rows),
                       effects = list(spat_id = 1:spde$n.spde), tag='loc') 

## estimate phylogenetic random effect for each species
stck.phy <- inla.stack(data = list(resp = NA), A = list(1),
                       effects=list(phy_id = poa_traits$phy_id), tag='phy') 

big_stack <- inla.stack(stack, stck.loc, stck.phy)
```


```{r}
full_mod <- inla(resp ~ 0 + intercept + Endophyte +
                   f(spat_id, model = spde) + ## spatial random effect
                   f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                 family = "gaussian",
                 data = inla.stack.data(big_stack, spde = spde),
                 control.predictor = list(A = inla.stack.A(big_stack), compute = TRUE,
                                          link = 1), 
                 control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE),
                 num.threads = 10 
)

summary(full_mod)
```


```{r}
grid <- inla.mesh.projector(spatial_mesh2, dims = c(501,501))


prd_mean <- inla.mesh.project(grid, full_mod$summary.ran$spat_id$mean) 
prd_sd <- inla.mesh.project(grid, full_mod$summary.ran$spat_id$sd)

grid_df <- tibble(long = grid$lattice$loc[ , 1], 
                  lat = grid$lattice$loc[ , 2],
                  prediction_mean = as.vector(prd_mean), 
                  prediction_sd = as.vector(prd_sd))
```


```{r}
bbox <- st_bbox(coast, crs = "+proj=longlat +datum=WGS84 +no_defs")

# Convert to sf polygon
grid_sf <- st_as_sf(grid_df, coords = c("long", "lat"), crs = 4326)

land <- st_union(coast)

grid_crop <- grid_sf[st_intersects(grid_sf, land, sparse = FALSE), ]

grid_crop <- grid_crop %>%
  mutate(long = st_coordinates(grid_crop)[,1],
         lat  = st_coordinates(grid_crop)[,2])

spat_map <- ggplot(coast) +
  geom_sf() +
  geom_raster(data = grid_crop, aes(x = long, y = lat, fill = prediction_mean)) + 
  # geom_sf(data = sea, fill = "white") +
  scale_fill_viridis(name = "Spatial effect", na.value = NA) + 
  theme_map() 

spat_map
```





Mesh 3
```{r}
coords <- st_as_sf(occ_data, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)


boundary <- list(inla.nonconvex.hull(coords, 2, resolution=c(200,200), crs=4326),
                 inla.nonconvex.hull(coords, 8, resolution=c(100,100), crs=4326 ))

```

```{r}
## Build the mesh:
spatial_mesh3 <- inla.mesh.2d(boundary=boundary, 
                             max.edge=c(1, 8),
                             cutoff= 1, ## Filter away adjacent points. 
                             offset=c(1, 2))



plot(spatial_mesh3)
plot(spatial_mesh3); points(occ_data$decimalLongitude, occ_data$decimalLatitude, col="red", cex = 0.1)
spatial_mesh3$n
```

setting the priors for the spatial effect
```{r}
spde <- inla.spde2.matern(mesh = spatial_mesh3, alpha = 2)
```


```{r}
A <- inla.spde.make.A(spatial_mesh3, loc = coords)
dim(A)

row_counts <- rowSums(A != 0); head(row_counts)
row_sums <- rowSums(A); head(row_sums)
# apply(A, 1, function(x) x[x > 0]) %>% t %>% head
```


```{r}
specs <- poa_traits$species

#Loop over species and compute one A-row per species
spec_A_list <- lapply(specs, function(x) {
  rows_x <- which(poa_traits_occ$species == x)
  A_x <- A[rows_x, , drop = FALSE]
  A_x_mean <- apply(A_x, 2, mean)

  return(A_x_mean)
})

spec_A_rows <- do.call(rbind, spec_A_list)


## convert back to sparse matrix
spec_A_rows <- Matrix(spec_A_rows, sparse = TRUE) 

dim(spec_A_rows)
apply(spec_A_rows, 1, sum) %>% head

apply(spec_A_rows, 1, function(x) sum(x > 0)) 

```




```{r}
stack <- inla.stack(data = list(resp = poa_traits$log_maxss.range.size.sqkm_st),
                    A = list(spec_A_rows, 1, 1, 1), 
                    effects=list(spat_id = 1:spde$n.spde,
                                 phy_id = poa_traits$phy_id,
                                 intercept = rep(1, nrow(poa_traits)), 
                                 list(Endophyte = poa_traits$Endophyte_status
                                      )), tag='est') 
```


```{r}
## estimate spatial random effect for each species
stck.loc <- inla.stack(data = list(resp = NA),
                       A = list(spec_A_rows),
                       effects = list(spat_id = 1:spde$n.spde), tag='loc') 

## estimate phylogenetic random effect for each species
stck.phy <- inla.stack(data = list(resp = NA), A = list(1),
                       effects=list(phy_id = poa_traits$phy_id), tag='phy') 

big_stack <- inla.stack(stack, stck.loc, stck.phy)
```



```{r}
full_mod <- inla(resp ~ 0 + intercept + Endophyte +
                   f(spat_id, model = spde) + ## spatial random effect
                   f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                 family = "gaussian",
                 data = inla.stack.data(big_stack, spde = spde),
                 control.predictor = list(A = inla.stack.A(big_stack), compute = TRUE,
                                          link = 1), 
                 control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE),
                 num.threads = 10 
)

summary(full_mod)
```


```{r}
grid <- inla.mesh.projector(spatial_mesh3, dims = c(501,501))


prd_mean <- inla.mesh.project(grid, full_mod$summary.ran$spat_id$mean) 
prd_sd <- inla.mesh.project(grid, full_mod$summary.ran$spat_id$sd)

grid_df <- tibble(long = grid$lattice$loc[ , 1], 
                  lat = grid$lattice$loc[ , 2],
                  prediction_mean = as.vector(prd_mean), 
                  prediction_sd = as.vector(prd_sd))
```


```{r}
bbox <- st_bbox(coast, crs = "+proj=longlat +datum=WGS84 +no_defs")

# Convert to sf polygon
grid_sf <- st_as_sf(grid_df, coords = c("long", "lat"), crs = 4326)

land <- st_union(coast)

grid_crop <- grid_sf[st_intersects(grid_sf, land, sparse = FALSE), ]

grid_crop <- grid_crop %>%
  mutate(long = st_coordinates(grid_crop)[,1],
         lat  = st_coordinates(grid_crop)[,2])

spat_map <- ggplot(coast) +
  geom_sf() +
  geom_raster(data = grid_crop, aes(x = long, y = lat, fill = prediction_mean)) + 
  # geom_sf(data = sea, fill = "white") +
  scale_fill_viridis(name = "Spatial effect", na.value = NA) + 
  theme_map() 

spat_map
```









now to try some different spatial priors, using spatial mesh 1 because this was the best one

Mesh1
```{r}
coords <- st_as_sf(occ_data, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)


boundary <- list(inla.nonconvex.hull(coords, 5, resolution=c(60,60), crs=4326),
                 inla.nonconvex.hull(coords, 12, resolution=c(50,50), crs=4326 ))

```

```{r}
## Build the mesh:
spatial_mesh1 <- inla.mesh.2d(boundary=boundary, 
                             max.edge=c(5, 12),
                             # min.angle=30,
                             cutoff= 3, ## Filter away adjacent points. 
                             offset=c(1, 2))



plot(spatial_mesh1)
plot(spatial_mesh1); points(occ_data$decimalLongitude, occ_data$decimalLatitude, col="red", cex = 0.1)
spatial_mesh1$n
```


```{r}
A <- inla.spde.make.A(spatial_mesh1, loc = coords)
dim(A)

row_counts <- rowSums(A != 0); head(row_counts)
row_sums <- rowSums(A); head(row_sums)
apply(A, 1, function(x) x[x > 0]) %>% t %>% head
```


```{r}
specs <- poa_traits$species

# Loop over species and compute one A-row per species
spec_A_list <- lapply(specs, function(x) {
  rows_x <- which(poa_traits_occ$species == x)
  A_x <- A[rows_x, , drop = FALSE]
  A_x_mean <- apply(A_x, 2, mean)

  return(A_x_mean)
})


# Combine species rows into a single matrix
spec_A_rows <- do.call(rbind, spec_A_list)

## convert back to sparse matrix
spec_A_rows <- Matrix(spec_A_rows, sparse = TRUE) 

dim(spec_A_rows)
apply(spec_A_rows, 1, sum) %>% head

apply(spec_A_rows, 1, function(x) sum(x > 0)) 

```



```{r}
spde <- inla.spde2.pcmatern(
  mesh = spatial_mesh1, alpha = 2, 
  prior.range = c(2, 0.1), 
  prior.sigma = c(1, 0.1),
  constr = TRUE)
```


```{r}
stack <- inla.stack(data = list(resp = poa_traits$log_maxss.range.size.sqkm_st),
                    A = list(spec_A_rows, 1, 1, 1), 
                    effects=list(spat_id = 1:spde$n.spde,
                                 phy_id = poa_traits$phy_id,
                                 intercept = rep(1, nrow(poa_traits)), 
                                 list(Endophyte = poa_traits$Endophyte_status
                                      )), tag='est') 
```


```{r}
## estimate spatial random effect for each species
stck.loc <- inla.stack(data = list(resp = NA),
                       A = list(spec_A_rows),
                       effects = list(spat_id = 1:spde$n.spde), tag='loc') 

## estimate phylogenetic random effect for each species
stck.phy <- inla.stack(data = list(resp = NA), A = list(1),
                       effects=list(phy_id = poa_traits$phy_id), tag='phy') 

big_stack <- inla.stack(stack, stck.loc, stck.phy)
```



```{r}
full_mod <- inla(resp ~ 0 + intercept + Endophyte +
                   f(spat_id, model = spde) + ## spatial random effect
                   f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                 family = "gaussian",
                 data = inla.stack.data(big_stack, spde = spde),
                 control.predictor = list(A = inla.stack.A(big_stack), compute = TRUE,
                                          link = 1), 
                 control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE),
                 num.threads = 10 
)

summary(full_mod)
```









```{r}
spde <- inla.spde2.pcmatern(
  mesh = spatial_mesh1, alpha = 2,
  prior.range = c(50, 0.5),   # median practical range ~50 units
  prior.sigma = c(2, 0.5),
  constr = TRUE)
```


```{r}
stack <- inla.stack(data = list(resp = poa_traits$log_maxss.range.size.sqkm_st),
                    A = list(spec_A_rows, 1, 1, 1), 
                    effects=list(spat_id = 1:spde$n.spde,
                                 phy_id = poa_traits$phy_id,
                                 intercept = rep(1, nrow(poa_traits)), 
                                 list(Endophyte = poa_traits$Endophyte_status
                                      )), tag='est') 
```


```{r}
## estimate spatial random effect for each species
stck.loc <- inla.stack(data = list(resp = NA),
                       A = list(spec_A_rows),
                       effects = list(spat_id = 1:spde$n.spde), tag='loc') 

## estimate phylogenetic random effect for each species
stck.phy <- inla.stack(data = list(resp = NA), A = list(1),
                       effects=list(phy_id = poa_traits$phy_id), tag='phy') 

big_stack <- inla.stack(stack, stck.loc, stck.phy)
```



```{r}
full_mod <- inla(resp ~ 0 + intercept + Endophyte +
                   f(spat_id, model = spde) + ## spatial random effect
                   f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                 family = "gaussian",
                 data = inla.stack.data(big_stack, spde = spde),
                 control.predictor = list(A = inla.stack.A(big_stack), compute = TRUE,
                                          link = 1), 
                 control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE),
                 num.threads = 10 
)

summary(full_mod)
```









```{r}
spde <- inla.spde2.pcmatern(
  mesh = spatial_mesh1, alpha = 2, ### mesh and smoothness parameter
  prior.range = c(20, 0.1), ### P(practic.range<2)=0.1
  prior.sigma = c(1, 0.1),
  constr = TRUE)
```


```{r}
stack <- inla.stack(data = list(resp = poa_traits$log_maxss.range.size.sqkm_st),
                    A = list(spec_A_rows, 1, 1, 1), 
                    effects=list(spat_id = 1:spde$n.spde,
                                 phy_id = poa_traits$phy_id,
                                 intercept = rep(1, nrow(poa_traits)), 
                                 list(Endophyte = poa_traits$Endophyte_status
                                      )), tag='est') 
```


```{r}
## estimate spatial random effect for each species
stck.loc <- inla.stack(data = list(resp = NA),
                       A = list(spec_A_rows),
                       effects = list(spat_id = 1:spde$n.spde), tag='loc') 

## estimate phylogenetic random effect for each species
stck.phy <- inla.stack(data = list(resp = NA), A = list(1),
                       effects=list(phy_id = poa_traits$phy_id), tag='phy') 

big_stack <- inla.stack(stack, stck.loc, stck.phy)
```


```{r}
full_mod <- inla(resp ~ 0 + intercept + Endophyte +
                   f(spat_id, model = spde) + ## spatial random effect
                   f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                 family = "gaussian",
                 data = inla.stack.data(big_stack, spde = spde),
                 control.predictor = list(A = inla.stack.A(big_stack), compute = TRUE,
                                          link = 1), 
                 control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE),
                 num.threads = 10 
)

summary(full_mod)
```







```{r}
spde <- inla.spde2.pcmatern(
  mesh = spatial_mesh1, alpha = 2,
  prior.range = c(5, 0.5),   
  prior.sigma = c(0.3, 0.5),
  constr = TRUE)
```


```{r}
stack <- inla.stack(data = list(resp = poa_traits$log_maxss.range.size.sqkm_st),
                    A = list(spec_A_rows, 1, 1, 1), 
                    effects=list(spat_id = 1:spde$n.spde,
                                 phy_id = poa_traits$phy_id,
                                 intercept = rep(1, nrow(poa_traits)), 
                                 list(Endophyte = poa_traits$Endophyte_status
                                      )), tag='est') 
```


```{r}
## estimate spatial random effect for each species
stck.loc <- inla.stack(data = list(resp = NA),
                       A = list(spec_A_rows),
                       effects = list(spat_id = 1:spde$n.spde), tag='loc') 

## estimate phylogenetic random effect for each species
stck.phy <- inla.stack(data = list(resp = NA), A = list(1),
                       effects=list(phy_id = poa_traits$phy_id), tag='phy') 

big_stack <- inla.stack(stack, stck.loc, stck.phy)
```


```{r}
full_mod <- inla(resp ~ 0 + intercept + Endophyte +
                   f(spat_id, model = spde) + ## spatial random effect
                   f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                 family = "gaussian",
                 data = inla.stack.data(big_stack, spde = spde),
                 control.predictor = list(A = inla.stack.A(big_stack), compute = TRUE,
                                          link = 1), 
                 control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE),
                 num.threads = 10 
)

summary(full_mod)
```






```{r}
spde <- inla.spde2.pcmatern(
  mesh = spatial_mesh1, alpha = 2,
  prior.range = c(3, 0.5),   # P(range < 5) = 0.5  (centred around posterior)
  prior.sigma = c(0.3, 0.5),
  constr = TRUE
)
```


```{r}
stack <- inla.stack(data = list(resp = poa_traits$log_maxss.range.size.sqkm_st),
                    A = list(spec_A_rows, 1, 1, 1), 
                    effects=list(spat_id = 1:spde$n.spde,
                                 phy_id = poa_traits$phy_id,
                                 intercept = rep(1, nrow(poa_traits)), 
                                 list(Endophyte = poa_traits$Endophyte_status
                                      )), tag='est') 
```


```{r}
## estimate spatial random effect for each species
stck.loc <- inla.stack(data = list(resp = NA),
                       A = list(spec_A_rows),
                       effects = list(spat_id = 1:spde$n.spde), tag='loc') 

## estimate phylogenetic random effect for each species
stck.phy <- inla.stack(data = list(resp = NA), A = list(1),
                       effects=list(phy_id = poa_traits$phy_id), tag='phy') 

big_stack <- inla.stack(stack, stck.loc, stck.phy)
```


```{r}
full_mod <- inla(resp ~ 0 + intercept + Endophyte +
                   f(spat_id, model = spde) + ## spatial random effect
                   f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                 family = "gaussian",
                 data = inla.stack.data(big_stack, spde = spde),
                 control.predictor = list(A = inla.stack.A(big_stack), compute = TRUE,
                                          link = 1), 
                 control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE),
                 num.threads = 10 
)

summary(full_mod)
```









```{r}
spde <- inla.spde2.pcmatern(
  mesh = spatial_mesh1, alpha = 2,
  prior.range = c(30, 0.5),
  prior.sigma = c(1.5, 0.5),
  constr = TRUE
)
```


```{r}
stack <- inla.stack(data = list(resp = poa_traits$log_maxss.range.size.sqkm_st),
                    A = list(spec_A_rows, 1, 1, 1), 
                    effects=list(spat_id = 1:spde$n.spde,
                                 phy_id = poa_traits$phy_id,
                                 intercept = rep(1, nrow(poa_traits)), 
                                 list(Endophyte = poa_traits$Endophyte_status
                                      )), tag='est') 
```


```{r}
## estimate spatial random effect for each species
stck.loc <- inla.stack(data = list(resp = NA),
                       A = list(spec_A_rows),
                       effects = list(spat_id = 1:spde$n.spde), tag='loc') 

## estimate phylogenetic random effect for each species
stck.phy <- inla.stack(data = list(resp = NA), A = list(1),
                       effects=list(phy_id = poa_traits$phy_id), tag='phy') 

big_stack <- inla.stack(stack, stck.loc, stck.phy)
```


```{r}
full_mod <- inla(resp ~ 0 + intercept + Endophyte +
                   f(spat_id, model = spde) + ## spatial random effect
                   f(phy_id, model = "generic0", Cmatrix = phylo_prec_mat, constr = TRUE, hyper = pcprior),
                 family = "gaussian",
                 data = inla.stack.data(big_stack, spde = spde),
                 control.predictor = list(A = inla.stack.A(big_stack), compute = TRUE,
                                          link = 1), 
                 control.compute = list(dic = TRUE, waic = TRUE, cpo = TRUE, config = TRUE),
                 num.threads = 10 
)

summary(full_mod)
```







